Reformulating phylogenetic mixed models to improve flexibility and speed

### 2018-11-19 phyloglmm_ms.tex
working on intro ...

Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or remote sensing of environmental variables), in combination with large-scale synthesis of existing data bases on species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
In response, researchers are studying more complicated systems and fitting more models to larger data sets describing species occurrence and traits.
One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting}. 
Comparative analyses explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; researchers may be interested in controlling for the statistical effects of phylogenetic correlation, quantifying the evolutionary signals that underly traits or both.
Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable, in phylogenetic regression, the evolutionary relationship between species are unobservable and assumed to evolve according to some underlying (latent) evolutionary process \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications such as biodiversity conservation, effects of climate change, and many more \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}. 
%such as ecological conser, the effects of climate change  \citep{santamaria2012evolution}. \mli{cite more here}.
While there exist a range of statistical tools for comparative analyses, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
In such cases, researchers may try to find ways to simplify their analyses (i.e. controlling for individual-varation of species and neglecting phylogenetic correlation between species \citep{bunnefeld2012island}, comparing species family \citep{tella1999habitat}, neglecting within-species variation \citep{ord2010adaptation}, and etc).
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
% \bmb{nice!}

@@

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
The typical linear mixed model has the form:
\begin{align}
\bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
g(\bmu) & = \bX \bbeta + \bZ \bb + \be \label{eq:glmm2} \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
\be & \sim \textrm{MVN}(0, \sigma^2_{residual} \label{eq:glmm3})
\end{align}
where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals.
Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.

>>>>

Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
Researchers are exploiting these data to fit complex models describing species occurrence and traits.
In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is
becoming ever more popular \bmb{might be interesting to some bibliometry, although absolutely not necessary \ldots}
PCAs explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; they can be used to control for the statistical effects of phylogenetic correlation, to quantify the phylogenetic signals in trait distributions, or both.
Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications such as biodiversity conservation, effects of climate change, and many more \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}. 
%such as ecological conser, the effects of climate change  \citep{santamaria2012evolution}. \mli{cite more here}.
While there exist a range of statistical tools for comparative analyses, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
In such cases, researchers may try to find ways to simplify their analyses (i.e. controlling for individual-varation of species and neglecting phylogenetic correlation between species \citep{bunnefeld2012island}, comparing species family \citep{tella1999habitat}, neglecting within-species variation \citep{ord2010adaptation}, and etc).
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
% \bmb{nice!}

@@

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
The typical linear mixed model has the form:
\begin{align}
\bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
g(\bmu) & = \bX \bbeta + \bZ \bb + \be \label{eq:glmm2} \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
\be & \sim \textrm{MVN}(0, \sigma^2_{residual} \label{eq:glmm4})
\end{align}
where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals.
Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.
